### 2019-03-28 Thesis/games.tex
Nightly commit.

The definition of the relation is in Figure \ref{FigIaSsOpSem}.
As an auxiliary definition, we need the notion of an \emph{evaluation context} (see \cite{Felleisen}).
This is a single-holed context defined inductively by the following BNF formula, where $M$ ranges over IA terms.
\begin{center}
\parbox{0.7\textwidth}{
\begin{mathpar}
  E \Coloneqq - \mid \suc E \mid \pred E \mid E;M \mid \If E\Then M \Else M \mid \IfO E \Then M \Else M \mid !E \mid M \gets E \mid E \gets n
\end{mathpar}}
\end{center}

\begin{figure}
  \begin{mathpar}
    \inferrule*{\Gamma,s\ts M \opto \Gamma',s',M'}{\Gamma,s\ts E[M] \opto \Gamma',s'\ts E[M']}
    \\
    \inferrule*{ }{\Gamma,s\ts (\lambda x.M)N \opto \Gamma,s\ts M[N/x]}
    \and
    \inferrule*{ }{\Gamma,s\ts \Y M \opto \Gamma,s\ts M(\Y M)}
    \and
    \inferrule*{ }{\Gamma,s\ts \succ n \opto \Gamma,s\ts n+1}
    \and
    \inferrule*{ }{\Gamma,s \ts \pred n+1 \opto \Gamma,s\ts n}
    \and
    \inferrule*{ }{\Gamma,s\ts \pred 0 \opto \Gamma,s\ts 0}
    \and
    \inferrule*{ }{\Gamma,s\ts \skipp;M \opto \Gamma,s\ts M}
    \and
    \inferrule*{ }{\Gamma,s\ts \If \true \Then N \Else P \opto \Gamma,s\ts N}
    \and
    \inferrule*{ }{\Gamma,s\ts \If \false \Then N \Else P \opto \Gamma,s\ts P}
    \and
    \inferrule*{ }{\Gamma,s\ts \IfO 0 \Then N \Else P \opto \Gamma,s\ts N}
    \and
    \inferrule*{ }{\Gamma,s\ts \IfO n+1 \Then N \Else P \opto \Gamma,s\ts P}
    \and
    \inferrule*{ }{\Gamma,s\ts x \gets n \opto \Gamma,(s\vert x\mapsto n)\ts \skipp}
    \and
    \inferrule*[right={$s(x)=n$}]{ }{\Gamma,s\ts \oc x \opto \Gamma,s\ts n}
    \\
    \inferrule*{ }{\Gamma,s\ts \neww \lambda x.M \opto \Gamma,x,(s\vert x\mapsto 0)\ts M}
    \\
    \inferrule*{ }{\Gamma,s \ts \mkvar W\,R \gets n \opto \Gamma,s\ts Wn}
    \and
    \inferrule*{ }{\Gamma,s \ts \oc\mkvar W\,R \opto \Gamma,s\ts R}
  \end{mathpar}
  \caption{Felleisen-style small-step operational semantics for Idealized Algol.}
  \label{FigIaSsOpSem}
\end{figure}

@@

\begin{proposition}
  Let $\Gamma$ be a $\Var$-context, let $s$ be a $\Gamma$-store, and let $\Gamma\ts M$ be an Idealized Algol term-in-context.  
  If $s'$ is another $\Gamma$-store and $\Gamma\ts c$ is a canonical form, then $\Gamma,s\ts M\converges c,s'$ if and only if there is some sequence
  \[
    \Gamma,s\ts M \opto \Gamma_1,s^{(1)},M_1 \opto \cdots \opto \Gamma_n,s^{(n)},M_n \opto \Gamma'',s'',c\,,
    \]
  where $\Gamma_1,\cdots,\Gamma_n,\Gamma''$ are $\Var$-contexts, $s^{(i)}$ is a $\Gamma_i$-store, $s''$ is a $\Gamma''$-store, $\Gamma_i\ts M_i$ are Idealized Algol terms in context, and $s''\vert_\Gamma=s'$.
\end{proposition}
\begin{proof}
  First, suppose that we have some sequence $\Gamma,s\ts M\opto\cdots\opto \Gamma'',s'',c$, where $s''\vert_\Gamma=\s'$.  
  We construct a derivation of $\Gamma,s\ts M\converges c,s'$ by induction on the length of this sequence.

  If the sequence has length $0$, then we can use the rule for the canonical form; i.e.
  \[
    \inferrule*{ }{\Gamma,s\ts c\converges c,\skipp}\,.
    \]
  Otherwise, suppose that the sequence starts $\Gamma,s\ts M\converges \Gamma_1,s^{(1)}\ts M_1$, and that we already have a derivation of
  \[
    \Gamma_1,s^{(1)} \ts M_1 \opto c,s''\vert_{\Gamma_1}\,.
    \]
  We split into cases based on the rule used for $\Gamma,s\ts M\converges \Gamma_1,s^{(1)}$.

\end{proof}

@@

>>>>

As an auxiliary definition, we need the notion of an \emph{evaluation context} (see \cite{Felleisen}).
This is a single-holed context defined inductively by the following BNF formula, where $M$ ranges over IA terms.
\begin{center}
\parbox{0.8\textwidth}{
\begin{mathpar}
  \EE \Coloneqq - \mid \EE\,M \mid \suc \EE \mid \pred \EE \mid \EE;M \mid \If \EE\Then M \Else M \mid \IfO \EE \Then M \Else M \mid !\EE \mid M \gets \EE \mid \EE \gets n
\end{mathpar}}
\end{center}

Next, we define a relation $\Gamma,s,M\oopto\Gamma',s',M'$ as in Figure \ref{FigIaSsOpSem}.

We then define the relation $\opto$ as
\[
  \inferrule*{ \Gamma,s\ts M\oopto \Gamma',s'\ts M'}{\Gamma,s\ts \EE[M] \opto \EE[M']}
  \]
for each evaluation context $\EE$.

\begin{figure}
  \begin{mathpar}
    \Gamma,s\ts (\lambda x.M)N \oopto \Gamma,s\ts M[N/x]
    \and
    \Gamma,s\ts \Y M \oopto \Gamma,s\ts M(\Y M)
    \and
    \Gamma,s\ts \suc n \oopto \Gamma,s\ts n+1
    \\
    \Gamma,s \ts \pred n+1 \oopto \Gamma,s\ts n
    \and
    \Gamma,s\ts \pred 0 \oopto \Gamma,s\ts 0
    \\
    \Gamma,s\ts \skipp;M \oopto \Gamma,s\ts M
    \\
    \Gamma,s\ts \If \true \Then N \Else P \oopto \Gamma,s\ts N
    \and
    \Gamma,s\ts \If \false \Then N \Else P \oopto \Gamma,s\ts P
    \and
    \Gamma,s\ts \IfO 0 \Then N \Else P \oopto \Gamma,s\ts N
    \and
    \Gamma,s\ts \IfO n+1 \Then N \Else P \oopto \Gamma,s\ts P
    \and
    \Gamma,s\ts x \gets n \oopto \Gamma,(s\vert x\mapsto n)\ts \skipp
    \and
    \Gamma,s\ts \oc x \oopto \Gamma,s\ts s(x)
    \\
    \Gamma,s\ts \neww \lambda x.M \oopto \Gamma,x,(s\vert x\mapsto 0)\ts M
    \\
    \Gamma,s \ts (\mkvar W\,R) \gets n \oopto \Gamma,s\ts Wn
    \and
    \Gamma,s \ts \oc(\mkvar W\,R) \oopto \Gamma,s\ts R
  \end{mathpar}
  \caption{Felleisen-style small-step operational semantics for Idealized Algol.}
  \label{FigIaSsOpSem}
\end{figure}

@@

\begin{lemma}
  Suppose that $\Gamma,s\ts M \oopto \Gamma',s'\ts M'$ and that $\Gamma',s'\ts \EE[M'] \converges c,s''$.  
  Then $\Gamma,s\ts \EE[M]\converges c,s''\vert_\Gamma$.
  \label{LemSmallToBig}
\end{lemma}
\begin{proof}
  Structural induction on $\EE$.
  The base case, when $\EE$ is a hole, covers the interesting cases, so we shall leave it to last.
  The remaining cases are quite similar, so we will show the proof for one of them for illustration.

  If $\EE=\EE'\,N$, for some term $N$, then we have $\Gamma',s'\ts \EE'[M']\,N\converges c,s''$.  
  By inspection of the rules in Figure \ref{FigIaOpSem}, the derivation of this must end with a rule of the form
  \[
    \inferrule*{\Gamma',s'\ts \EE'[M'] \converges \lambda x.M'',t \\ \Gamma',t \ts M''[N/x] \converges c,s''}
    {\Gamma',s'\ts \EE'[M']\,N \converges c,s''}\,.
    \]
  Thus, $\Gamma',s'\ts \EE'[M']\converges \lambda x.M'',t$ must be provable for some $M'',t$.
  By the induction hypothesis, this means that $\Gamma,s\ts \EE'[M]\converges \lambda x.M'',t$ is provable.
  Then we have a derivation
  \[
    \inferrule*{\Gamma',s\ts \EE'[M] \converges \lambda x.M'',t \\ \Gamma',t \ts M''[N/x] \converges c,s''}
    {\Gamma',s\ts \EE'[M]\,N\converges c,s''\vert_{\Gamma'}}\,.
    \]
  Then, because any extra variables in $\Gamma'$ are not mentioned in $\EE'[M]\,N$ or in $c$, we have
  \[
    \Gamma,s\ts \EE'[M]\,N\converges c,s''\vert_\Gamma\,.
    \]
  Now, let us suppose that $\EE$ is a hole, so that $\EE[M']=M'$.

  Then there are a number of cases, depending on the particular $\oopto$ rule we are using.  
  Many of these cases are similar, so we will cover a few of them for the purposes of illustration.
  \begin{itemize}
    \item Suppose that $\Gamma,s\ts M[N/x]\converges c,s'$.  
      Then we have a derivation
      \[
        \inferrule*
        {
          \inferrule*
          {
          }
          {
            \Gamma,s \ts \lambda x.M \converges \lambda x.M,s
          }
          \\
          \Gamma,s\ts M[N/x] \converges c,s'
        }
        {
          \Gamma,s\ts (\lambda x.M)N \converges c,s'
        }\,.
        \]
    \item We have a derivation
      \[
        \inferrule*
        {
          \inferrule*
          {
          }
          {
            \Gamma,s\ts x\converges x,s
          }
          \\
          \inferrule*
          {
          }
          {
            \Gamma,s\ts n\converges n,s
          }
        }
        {
          \Gamma,s\ts x\gets n\converges \skipp,(s\vert x\mapsto n)
        }\,.
        \]
    \item We have a derivation
      \[
        \inferrule*
        {
          \inferrule*
          {
          }
          {
            \Gamma,s\ts x \converges x,s'
          }
        }
        {
          \Gamma,s\ts \oc x\converges s(x),s
        }\,.
        \]
    \item Suppose that $\Gamma,x,(s\vert x\mapsto 0) \ts M\converges c,s'$.  
      Then we have a derivation
      \[
        \inferrule*
        {
          \Gamma,x,(s\vert x\mapsto 0) \ts M \converges c,s'
        }
        {
          \Gamma,s\ts \neww\lambda x.M \converges c,s'\vert_\Gamma
        }\,,
        \]
      since $s'=(s'\vert_\Gamma\vert x\mapsto s'(x))$.\qedhere
  \end{itemize}
\end{proof}

We have proved:

\begin{proposition}
  Suppose that we have a sequence
  \[
    \Gamma_1,s^{(1)},M_1\opto\cdots\opto\Gamma_n,s^{(n)},M_n\,,
    \]
  where $M_n$ is a canonical form.  
  Then $\Gamma_1,s^{(1)} \ts M_1\converges M_n,s^{(n)}\vert_{\Gamma_1}$.  
  \label{PropSmallToBig}
\end{proposition}
\begin{proof}
  Induction on $n$.  
  The inductive step is Lemma \ref{LemSmallToBig}, while the base case ($n=1$) is given by the derivation
  \[
    \inferrule*{ }
    {\Gamma,s\ts c\converges c,s}\,.\qedhere
    \]
\end{proof}

We can also prove the converse.

\begin{proposition}
  Suppose that $\Gamma,s\ts M\converges c,s'$.  
  Then there are sequences $\Gamma=\Gamma_1,\cdots,\Gamma_n$, $s=s^{(1)},\cdots,s^{(n)}$, $M=M_1,\cdots,M_n=c$ such that
  \[
    \Gamma_1,s^{(1)}\ts M_1 \opto \cdots \opto \Gamma_n,s^{(n)}\ts M_n\,,
    \]
  and $s^{(n)}\vert_{\Gamma}=s'$.
  \label{PropBigToSmall}
\end{proposition}
\begin{proof}
  Induction on the derivation of $\Gamma,s\ts M\converges c,s'$.  
  Since most of the cases are similar, we cover a selection for illustration.

  \begin{itemize}
    \item Suppose that the last step in the derivation is
      \[
        \inferrule*{ }
        {\Gamma,s\ts c\converges c,s}\,.
        \]
      Then we have the one-element sequence $\Gamma,s\ts c$.  
    \item Suppose that the last step in the derivation is
      \[
        \inferrule*
        {
          \Gamma,s\ts M\converges \lambda x.M',s' \\ \Gamma,s' \ts M'[N/x]\converges c,s''
        }
        {
          \Gamma,s \ts MN\converges c,s''
        }\,.
        \]
      Then, by the inductive hypothesis, we have small-step derivations
      \begin{mathpar}
        \Gamma,s \ts M \opto \cdots \opto \Gamma',t',\lambda x.M'
        \and
        \Gamma,s' \ts M'[N/x]\opto \cdots \opto \Gamma'',t'',c\,,
      \end{mathpar}
      where $t'\vert_\Gamma=s'$ and $t''\vert_\Gamma=s''$.

      If we apply the evaluation context $-N$ pointwise to the first small-step derivation, then we have another valid small-step derivation.  
      Then we can join the two together to get the derivation
      \begin{center}
        \parbox{0.8\textwidth}{
        \begin{mathpar}
          \Gamma,s \ts MN \opto \cdots \opto \Gamma',t',(\lambda x.M')N \opto \Gamma',t',M'[N/x] \opto \cdots \opto \Gamma'\cup\Gamma'',t''\setminus t',c\,,
        \end{mathpar}}
      \end{center}
      where $t''\setminus t'$ is the $\Gamma'\cup\Gamma''$-store that agrees with $t''$ on $\Gamma''$ and with $t'$ on $\Gamma'\setminus\Gamma''$.
      Then $(t''\setminus t')\vert_{\Gamma}=s''$.
    \item Suppose that the last step in the derivation is
      \[
        \inferrule*
        {
          \Gamma,s\ts E\converges n,s' \\ \Gamma,s'\ts V\converges x,s''
        }
        {
          \Gamma,s \ts V \gets E \converges\skipp,(s''\vert x\mapsto n)
        }\,.
        \]
      By the inductive hypothesis, we have small-step derivations
      \begin{mathpar}
        \Gamma,s\ts E \opto \cdots \opto \Gamma',t',n
        \and
        \Gamma,s'\ts V \opto \cdots \opto \Gamma'',t'',x\,,
      \end{mathpar}
      where $t'\vert_{\Gamma}=s'$ and $t''\vert_{\Gamma}=s''$.

      We may apply the evaluation context $V\gets-$ pointwise to the first derivation and the evaluation context $-\gets n$ pointwise to the second, and then string the two together to get
      \begin{center}
        \parbox{0.8\textwidth}{
        \begin{mathpar}
          \Gamma,s\ts V\gets E \opto \cdots \opto \Gamma',t',V\gets n \opto \cdots \opto \Gamma'\cup\Gamma'',t''\setminus t',x\gets n \opto \Gamma'\cup\Gamma'',(t''\setminus t'\vert x\mapsto n)\,,
        \end{mathpar}}
      \end{center}
      where we have $(t''\setminus t'\vert x\mapsto n)\vert_\Gamma=(s''\vert x\mapsto n)$.
    \item Suppose that the last step in the derivation is
      \[
        \inferrule*
        {
          \Gamma,s\ts V \converges x,s'
        }
        {
          \Gamma,s\ts \oc V \converges s'(x),s'
        }\,.
        \]
      Then, by the induction hypothesis, we have a small-step derivation
      \[
        \Gamma,s\ts V \opto \cdots \opto \Gamma',t'\ts x\,,
        \]
      where $t'\vert_\Gamma=s'$.  
      Then we may compose this derivation pointwise with the evaluation context $\oc-$, and add an extra term on the end, to arrive at the derivation
      \[
        \Gamma,s\ts \oc V \opto \cdots \opto \Gamma',t'\ts \oc x \opto \Gamma',t'\ts t'(x)\,,
        \]
      where $t'(x)=s'(x)$.
    \item Lastly, suppose that the last step in the derivation is
      \[
        \inferrule*
        {
          \Gamma,x,(s\vert x\mapsto 0) \ts M \converges c,(s'\vert x\mapsto n)
        }
        {
          \Gamma,s \ts \neww \lambda x.M \converges c,s'
        }\,.
        \]
      By the induction hypothesis, we have a small-step derivation
      \[
        \Gamma,x,(s\vert x\mapsto 0) \ts M \opto \cdots \opto \Gamma',x,(t'\vert x\mapsto n),c\,,
        \]
      where $t'\vert_\Gamma=s'$.  
      Then we may add a term at the beginning to give us
      \[
        \Gamma,s \ts \neww \lambda x.M \opto \Gamma,x,(s\vert x\mapsto 0) \ts M \opto \cdots \opto \Gamma',x,(t'\vert x\mapsto n),c\,,
        \]
      where $(t'\vert x\mapsto n) \vert_\Gamma=s'$.\qedhere
  \end{itemize}
\end{proof}

@@

To prove soundness of our model, we shall use the small-step formulation.  
Our reason for this is that the most difficult part of the denotational semantics we are using is the part to do with state.  
In the big-step formulation, nearly every rule involves the state changing in some way, whereas in the small-step formulation, only the rules that specifically pertain to the stateful components of the language do.

\begin{lemma}
  Suppose that $\Gamma,s\ts M \opto \Gamma',s'\ts N$.
\end{lemma}
